Monte Carlo estimation of the probability of causal contacts between communicating civilisations

### 2020-05-30 submit_v03/manuscript.tex
last day

\title[Probabilities of SETI contacts]{Probability of causal
contacts between interstellar civilizations through Monte Carlo simulations}
%
%A network of constrained causal contacts in the Galaxy as a proxy for
%the probability of 
%
% A model for constrained causal contacts between location in the GHZ

@@

%%% S E C T I O N - - - - - - - - - - - - - - - - - - - - - - -
\section{Introduction}\label{S_motivations}

The Drake equation  \citep{drake_intelligent_1962} provides a truly
helpful educated guess, a rational set of lenses --the factors in the
equation-- through which to look at future contacts with
technologically advanced civilizations in the Milky Way. 
%
The equation quantifies the number of civilizations from whom we might
receive an electromagnetic signal, using a collection of factors that
have been extensively discussed in the literature and whose estimates
are revised continually. 
%
A comprehensive review and analysis of each term of the equation are
presented in \citet{vakoch_drake_2015}.
%
Optimistic estimates from the Drake equation contrast with the
so-called Fermi paradox, which states the apparent contradiction between the
expected abundance of life in the Galaxy and the lack of evidence for
it \citep[e.g. ][]{hart_explanation_1975, brin_great_1983,
barlow_galactic_2012, forgan_galactic_2016, anchordoqui_is_2017,
Sotos_biotechnology_2019, carroll_nellemback_fermi_2019}.
%
There are many propositions aimed at solving this paradox, which make
use of statistical \citep{solomonides_probabilistic_2016, horvat_calculating_2007,
maccone_statistical_2015} or stochastic approaches
\citep{forgan_numerical_2009, bloetscher_using_2019,
glade_stochastic_2011, forgan_numerical_2010}.
%
Regarding the Drake equation, analytical interpretations
\citep{prantzos_joint_2013, smith_broadcasting_2009} or reformulations
\citep[][and references therein]{burchell_whither_2006} have also been
proposed. 
%
The absence of detections of extraterrestrial intelligent signals
could be explained by
astrophysical phenomena that makes life difficult to develop
\citep{annis_astrophysical_1999}.
%
Besides the possible scarcity of life,
alternative scenarios have also been discussed
\citep{barlow_galactic_2013, lampton_information_2013,
conway_three_2018, forgan_galactic_2017}.
%
The large distances in the Galaxy and the likely limited lifetime of
civilizations may play an important role in determining how difficult it
would be to obtain evidence for other inhabited worlds.
%
%We stress the fact that the lack of detections of extraterrestrial
%intelligent signals is the only observational fact that can be taken
%for sure.
The analysis of these scenarios is difficult due to the lack of data
about the hypothetical extraterrestrial intelligences.
%
Indeed, as \citep{tarter_search_2001} pointed out, according to our
current technical capabilities for the search of extraterrestrial
intelligence (SETI), we have not received any signal yet.
%
%Indeed, for the number of years projects for the Search of
%Extraterrestrial Intelligences (SETI) have been working, no signal has
%been received using the technical possibilities within their search
%criteria \citep{tarter_search_2001}.
%
The absence of detections has also motivated alternative ideas for new
SETI strategies \citep{forgan_exoplanet_2017, balbi_impact_2018,
loeb_eavesdropping_2006, maccone_KLT_2010, tarter_advancing_2009,
enriquez_breakthrough_2017, loeb_relative_2016, maccone_SETI_2011,
lingam_relative_2019, wright_theGsearch_2015, maccone_SETI_2013,
maccone_lognormals_2014, harp_application_2018,
forgan_possibility_2013, forgan_galactic_2017, funes_searching_2019}.

@@

The discussion about the problem of the (unknown)
abundance of civilisations in the Galaxy has been organised around the
factors in the Drake equation \citep{hinkel_interdisciplinary_2019}.
%
However, the uncertainties in these factors, specially the ones
representing biological processes, make it
less suitable to a formal study with the purpose of defining
searching strategies or
computing the estimated number of extraterrestrial intelligences.
%
%The optimistic estimations from the Drake equation also contrast with
%the results from the SETI initiatives.
%
%Notably, despite decades of effort not a single evidence for
%intelligent signals has been detected.
%
A number of studies propose alternative formalisms for the estimation of the
likelihood of detecting intelligent signals from space.
%
\citet{prantzos_joint_2013}, for example, proposes a unified framework for a joint
analysis of the Drake equation and the Fermi paradox, concluding that
for sufficiently long-lived civilisations, colonisation
is the most promising strategy to find other life forms.
%
\citet{haqq-misra_drake_2017} discuss the dependence of the Drake
equation parameters on the spectral type of the host stars and the
time since the Galaxy formed, and examine trajectories for the
emergence of communicative civilisations.
%
Some modifications to the original idea of the Drake equation have
been proposed, in order to
integrate a temporal structure, to
reformulate it as a stochastic process or to propose alternative
probabilistic expressions.
%
Temporal aspects of the distribution of communicating civilisations
and their contacts have been explored by several authors
\citep{fogg_temporal_1987, forgan_spatiotemporal_2011,
balbi_impact_2018, balb_spatiotemporal_2018, horvat_impact_2011}, as
well as efforts on considering the stochastic nature of the Drake
equation \citep{glade_stochastic_2011}.
%
The literature about this topic brings also a number of works that consider
numerical simulations \citep{forgan_evaluating_2015,
vukotic_grandeur_2016, murante_simulating_2015, forgan_numerical_2009,
forgan_galactic_2017, ramirez_new_2017}.
%
Although this approach does not rely on values that can be measured
from observations
(like the fraction of stars with planets), 
it depends on the definition of
unknown or uncertain parameters required to carry out
a the simulations.
%
\citet{grimaldi_signal_2017} considers a statistical model for the
probability of the Earth contacting other intelligent civilisations,
taking into account the finite lifetime of signal emitters, and based
on the fractional volume occupied by all signals reaching our planet.

@@

A notable example of statistica modeling is the work by
\citet{balbi_impact_2018}, who investigates the chance of
communicating civilisations making causal contact within a volume
surrounding the location of the Earth.
%
The author argues that the causal contact requirement involves mainly
the distance between civilisations, their lifespan and their times of
appearance.
%
This is important since the time the light takes to travel across the
Galaxy might be much lesser than the lifetime of the emitter.
%
\citet{balbi_impact_2018} fixes the total number of civilisations and
explores the parametric space that comprises three parameters, namely,
the distance to the Earth, the time of appearance and the lifespan of
the communicating civilization.
%
Each of these three variables are drawn from a random distribution.
%
The distribution of the distances results from a uniform model for the
positions of civilizations within the plane of the Galaxy.
%
For the distribution of the characteristic time of appearance, the
author explores exponential and truncated Gaussian distributions,
while for the distribution of the lifespans chooses an exponential
distribution.
%
It is important to consider that the estimation of the number of
communicating civilisations vary with the choice of the statistical
distribution for the time of appearance, as shown by
\citet{balbi_impact_2018}.
%
For all analysed distributions, the author concludes that, the
fraction of emitters that are listened is low if they are spreaded in
time and with limited lifetimes.
%
In another numerical approach, \citet{vukotic_astrobiological_2012}
propose a probabilistic cellular automata modelling. 
%
In this framework, a complex system is modelled by a lattice of cells
which evolve at discrete time steps, according to transition rules
that take into account the neighbour cells.
%
The authors implement this model to a network of cells which represent
life complexity on a 2-dimensional region resembling the Galactic
Habitable Zone (GHZ), an annular ring set between a minimal radius of
6 kpc and a peak radius of 10 kpc.
%
This simulations represent the spread if intelligence as an
implementation of panspermia theories.
%
Within this framework, \citet{vukotic_astrobiological_2012} also make
several Monte Carlo simulations and analyse ensemble-averaged results.
%
Their work aims at analysing the evolution of life, and although it
does not account for the network of causal contacts among
technological civilisations, it offers a tool to think SETI searches
from a novel point of view.

@@

This study is conceived as an introduction to a simple probabilistic
model and a numerical exploration of its parameter space.
%
With these tools, we address the problem of the temporal and spatial structure of
the distribution of communicating civilisations.
%
This approach does not require asumptions about, e.g., the origin of life, the development of
intelligence or the formation of habitable planets according to stellar type.
%
Instead, we assume monoparametric function families to model the appearance of
points in the disk of the Galaxy and over the time, which we call
\textit{nodes}.
%
These nodes represent the locations of ideal intelligent agents that are able
to receive and emit signals with perfect efficiency in all directions.
%
Then we analyse the network of causally connected nodes limited to have a
maximum separation representing the maximum distance a signal can travel with
an intensity above a fixed threshold.
%
A node is causally connected to other node if it is within their light cone.
%
We adopt as a definition that a light cone is the region in spacetime within
the surface generated by the light emanating from a given point in space in a
given period of time.
%
For example, two nodes separated by 100 light years which appear with a
difference in time of 80 years and last 50 years each, will be causally
connected for 30 years.
%
This connection is not, though, bidirectional.
%
In fact, the second node will see the first one for 30 years, but the first
node will never acknowledge the existence of the second.
%
This simple tought experiment exemplify the importance of the time variable to
analyse the structure of connections.
%
The model can be described with three parameters, namely, the mean separation
between the appearance of nodes, the mean lifetime of the nodes, and the
maximum separation to allow an effective causal contact.
%
The model presented here comprises the stochastic networks of constrained
causally connected nodes, together with the three parameters and additional
hypotheses.
%
Hereafter, we refeer to this model as the C3Net model, standing for
''Constrained Causally Conencted Network``.

We provide a framework to explore, through a suite of numerical simulations,
the parameter space of three unknown observables.
%
This allows to discuss possible scenarios and their consequences in terms of
the probability of making contacts.
%
The method we use for simulating a stochastic process is an approximation that
allows to study the behaviour of complex systems, by considering a sequence of
well defined discrete events.
%
The simulation is carried out by following all the variables that describe and
constitute the state of the system, and the evolution of the process is
described as a set of changes in that state.
%
In this context, an event produces a specific change in the state that can be
triggered by random variables that encode the stochastic nature of the physical
phenomenon.
%
For example, when a new contact is produced between two entities in the
simulated galaxy, the number of active communication lines is increased by one.
%
Also, if it is the first contact for that nodes, 
then the number of communicated nodes increases by two.
%
When a new node becomes active, the sistem has an increase of one
in the number of nodes, although the number of communcations does not
necessarily change.
%
The process then involves following the changes on the state of the system, defining
the initial and final states.
%
This is done by defining a method that allows to keep track of the time
progress in steps and maintaining a list of relevant events, i.e., the events
that produce a change in the vriables of interest.
%
With this method we abandon the frequentist approach of the Drake equation to
compute the number of civilisations, providing instead its statistical
distribution.
%
More importantly, we are interested on the probability of contacts, which
depend critically on the time variable.
%
This is an exploratory analysis that aims at developing a numerical tool to
discuss the different scenarios based on statistical heuristics.
%
The approach proposed here should be considered as a compromise between the
uncertainties of the frequentist estimations and the detailed recipes required
on the numerical simulations.
%
It is worth noticing that the model is not intended for a formal fit at this
stage due to the lack of data, but is can help to understand how unusual it
would be to actively search for intelligente signal for 50 years without
success.

This paper is organized as follows.
%
In Sec.~\ref{S_methods} we introduce the methods and discuss the candidate
distributions for the statistical aspects of the times involved in the
communication process.
%
We present our results in Sec~\ref{S_results}, with special emphasis on the
statistical distributions of the duration of causal contacts in one or both
directions and the distribution of time intervals for the waiting of the first
contact.
%
This quantities are considered as a function of the three simulation
parameters.
%
In Sec.~\ref{S_discussion} we discuss our results and future research
directions.   
%
In this work we use the International System of units recommended by the
International Astronomial Union, where the simbol for years is ''a''.

\begin{figure}[!t]
   %
   \centering
   %
   \includegraphics[width=0.5\textwidth]{F_scheme.pdf}
   %
   \caption{
   %
Space--time diagrams showing a schematic representation of the different stages
in the development of a constrained causal contact node (\ceti{}).
%
Panel (a) represents the region $R$ in space and time which is causally
connected to the emitter (E, left vertex), following an ``A'' type
event (``Awakening'') in which the node acquires the communication
capability.
%
The sphere of first contact (SFC) is the sphere centered on the emitter that
grows until its radius reaches the D$_{\rm max}$ distance, in which the
power of the signal would equal the detectability threshold.
%
In the Figure this sphere is represented by the left triangle of the region
$R$.
%
The surface of last contact (SLC) is another sphere that grows from a ``D''
event (``Doomsday'').
%
The region which is causally connected to the emitter is then limited by these
two spheres, and has the shape of a sphere or of a spherical shell,
depending on the time.
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b).
%
The receiver node E$_1$ can listen signals from the emitter node E$_2$, from
the ``Contact'' event (t=C$_{12}$) up to the ``Blackout'' event
(t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$. 
%
} \label{F_scheme}
%
\end{figure}

%%% S E C T I O N - - - - - - - - - - - - - - - - - - - - - - -
\section{Methods and working hypotheses}\label{S_methods}

Simulations are suitable tools to analyse systems that evolve with time and
involve randomness.
%
An advantage of a numerical approach is that it usually requires fewer assumptions
and simplifications, and can be applied to systems where analytical models are
hard or impossible to develop.
%
In particular, a suitable tool to model complex stochastic processes through
the changes in the state of a system is the discrete--event (hereafter, DE)
simulation approach.
%
A system described with the DE paradigm is characterised by a set of actors and
events.
%
Actors interact causally through a series of events on a timeline and process
them in chronological order \citep{ptolemaeus_system_2014,
chung_simulation_2003, ross_simulation_2012}.
%
Each event modify the variables that define the process, producing the
corresponding change in the state of the simulated system.
%
This method is well suited for the particular case of the diffusion of
intelligent signals in the galaxy and allows to explore several models easily.
%
We simulate the statistical properties of a set of points in space and time
that have a causal connection at light speed and are separated
by a maximum distance.
%
We refer to this nodes as ''Constrained Causal Contact Nodes``
(hereafter, \ceti{}).
%
We choose this generic name in order to stress the fact that in this analysis 
only the causal contact is considered, independently of any broadcasting or
lookout activity.
%
The system we propose is ideal, in the sense that it considers the special case
of a fully efficient node that emits and receives isotropically.
%
A causal contact node, considered as a broadcasting station that has the
ability to detect signals through an active search program, could have a
lesser than one efficiency factor, which is not included in the current
analysis.
%
Also, it is worth mentioning that this is a general approach, and not
necessarily a \ceti{} is the host of an intelligent civilization.
%
It can be associated with a planet where life has developed,
became intelligent, reached the skills required to find the right communication
channel, sustained a search and established a contact.
%
Alternative message processing entities could be considered, for example
interstellar beacons where life has ceased to exist but continue with its
emission, or communication stations established by probes or left by
intelligent beings \citep[see, e.g., ][]{peters_outer_2018,
barlow_galactic_2013}.
%
In principle, these strategies could affect our results since it would be
easier to configure a cluster of \cetis{} that spread in time.
%
However, we do not consider these speculative alternatives at this point.
%
For the purpose of the present analysis, only the communication capability is
relevant, since we study the causal contacts between the locations.
%
The system is defined by a number of actors that represent \cetis{} and appear
at different instants in time, generating events that produce meaningful
changes in the variables that describe the system, i.e., in the arrangement of
\cetis{} and their network of causal contacts.
%
For example, the appearance of a new \ceti{} in a region filled with a signal
emitted by another node, will increase the number of active \cetis{} and the
number of pairs of \cetis{} in causal contact.
%
Assuming some simple hypotheses, the discrete events method can be performed
taking into account a small number of variables, which allow to analyze the
variation of the results in the model parameter space.

@@

The temporal structure that emerges from the experimental setup implies that,
as a fundamental property of the simulations, a causal connection can be
produced without  requiring that the two nodes are active at the same time.
%
This property of the system arises as a consequence of the large spatial and
temporal scales, where a message is transmitted at the (relatively small) light
speed.
%
Although a \ceti{} could be active for a large enough period to transmit a
message at large distances, the limited power of the message and the dilution
that depends on the squared distance from the source imposes a detectability
limit.
%
As a consequence of this limitation and of their finite lifetime, considered as
the period between the acquisition and loss of communicating capacity, each
\ceti{} will fill a spherical shell region of the galaxy, limited by two
concentric spherical surfaces.
%
The leading front, or surface of first contact (SFC) grows from the central
\ceti{} until it reaches the maximum distance $D_{\rm max}$.
%
Following the end of the civilization, there is still a region which is filled
with the emitted signals.
%
This approach has been also considered in other statistical models \citep[e.g.,
][]{smith_broadcasting_2009, grimaldi_signal_2017, Grimaldi2018}.
%
The trailing front, or surface of last contact (SLC) also grows from the
central \ceti{}, with a delay with respect to the SFC equivalent to the
lifetime of the node, and produces a spherical shell region.
%
Any other \ceti{} within this region will be in causal contact with the originating node,
even if it has disappeared before the time of contact.
%
This region will grow if the surface of first contact has not yet reached the
maximum distance $D_{\rm max}$, and will shrink after a D-type event until the
surface of last contact reaches $D_{\rm max}$, producing as a result the loss
of all signal from the central \ceti{}.
%
In our approach, we consider a model galaxy where the width of the disk is
negligible with respect to the radius of the disk.
%
In the 2D simulation only the intersection of the communicating spherical
shells with the plane of the galaxy is relevant, and produce the corresponding
circles or rings for the filled spheres or annular regions of the spherical
shells, respectively.
%
The initially growing communicating sphere is shown over space--time diagrams,
where time is represented on the horizontal direction, and space is represented
in the vertical direction.
%
In the Fig.~\ref{F_scheme} the two emitters in panel (b), E$_1$ and E$_2$,
reach each other at different times.
%
The time span for E$_i$ is (A$_i$ , D$_i$) , for $i = {1, 2}$.
%
Emitter $i$ can listen to emitter $j$ between C$_{ij}$ and B$_{ij}$.
%
The type and length of causal contact in both directions depend on the distance
and time lag between the awakening events, the maximum distance that a signal
can reach and the time period in which each emitter is active.

@@

   $\tau_{a}$ & Mean temporal separation between consecutive
   awakenings & 4000 yr & 200\,000 yr & 50\\ 
   $\tau_{s}$ & Mean lifetime of a \ceti{}
   & 10000 yr & 500\,000 yr& 50\\ 
   D$_{max}$ & Maximum reach of a message  &  \multicolumn{2}{l}{3 kpc, 12 kpc, 25 kpc} & 3\\
\hline

>>>>

\title[Probability of causal contacts]{Monte Carlo estimation of the probability 
of causal contacts between communicating civilisations}

@@

\section{Introduction}\label{S_motivations}

The Drake equation  \citep{drake_intelligent_1962} provides a truly
helpful educated guess, a rational set of lenses --the factors in the
equation-- through which to look at future contacts with
technologically advanced civilizations in the Milky Way. 
%
The equation quantifies the number of civilizations from whom we might
receive an electromagnetic signal, using a collection of factors that
have been extensively discussed in the literature and whose estimates
are revised continually. 
%
A comprehensive review and analysis of each term of the equation are
presented in \citet{vakoch_drake_2015}.
%
Optimistic estimates from the Drake equation contrast with the
so-called Fermi paradox, which states the apparent contradiction between the
expected abundance of life in the Galaxy and the lack of evidence for
it \citep[e.g. ][]{hart_explanation_1975, brin_great_1983,
barlow_galactic_2012, forgan_galactic_2016, anchordoqui_is_2017,
Sotos_biotechnology_2019, carroll_nellemback_fermi_2019}.
%
There are many propositions aimed at solving this paradox, which make
use of statistical \citep{solomonides_probabilistic_2016, horvat_calculating_2007,
maccone_statistical_2015} or stochastic approaches
\citep{forgan_numerical_2009, bloetscher_using_2019,
glade_stochastic_2011, forgan_numerical_2010}.
%
Regarding the Drake equation, analytical interpretations
\citep{prantzos_joint_2013, smith_broadcasting_2009} or reformulations
\citep[][and references therein]{burchell_whither_2006} have also been
proposed. 
%
The absence of detections of extraterrestrial intelligent signals
could be explained by
astrophysical phenomena that makes life difficult to develop
\citep{annis_astrophysical_1999}.
%
Besides the possible scarcity of life,
alternative scenarios have also been discussed
\citep{barlow_galactic_2013, lampton_information_2013,
conway_three_2018, forgan_galactic_2017}.
%
The large distances in the Galaxy and the likely limited lifetime of
civilizations may play an important role in determining how difficult it
would be to obtain evidence for other inhabited worlds.
%
The analysis of these scenarios is difficult due to the lack of data
about the hypothetical extraterrestrial intelligences.
%
Indeed, as \citep{tarter_search_2001} pointed out, according to our
current technical capabilities for the search of extraterrestrial
intelligence (SETI), we have not received any signal yet.
%
The absence of detections has also motivated alternative ideas for new
SETI strategies \citep{forgan_exoplanet_2017, balbi_impact_2018,
loeb_eavesdropping_2006, maccone_KLT_2010, tarter_advancing_2009,
enriquez_breakthrough_2017, loeb_relative_2016, maccone_SETI_2011,
lingam_relative_2019, wright_theGsearch_2015, maccone_SETI_2013,
maccone_lognormals_2014, harp_application_2018,
forgan_possibility_2013, forgan_galactic_2017, funes_searching_2019}.

@@

The discussion about the problem of the unknown
abundance of civilisations in the Galaxy has been organised around the
factors in the Drake equation \citep{hinkel_interdisciplinary_2019}.
%
However, the uncertainties in these factors, specially the ones
representing biological processes, make it
less suitable to a formal study with the purpose of defining
searching strategies or
computing the estimated number of extraterrestrial intelligences.
%
A number of studies propose alternative formalisms for the estimation of the
likelihood of detecting intelligent signals from space.
%
\citet{prantzos_joint_2013}, for example, proposes a unified framework for a joint
analysis of the Drake equation and the Fermi paradox, concluding that
for sufficiently long-lived civilisations, colonisation
is the most promising strategy to find other life forms.
%
\citet{haqq-misra_drake_2017} discuss the dependence of the Drake
equation parameters on the spectral type of the host stars and the
time since the Galaxy formed, and examine trajectories for the
emergence of communicative civilisations.
%
Some modifications to the original idea of the Drake equation have
been proposed, in order to
integrate a temporal structure, to
reformulate it as a stochastic process or to propose alternative
probabilistic expressions.
%
Temporal aspects of the distribution of communicating civilisations
and their contacts have been explored by several authors
\citep{fogg_temporal_1987, forgan_spatiotemporal_2011,
balbi_impact_2018, balb_spatiotemporal_2018, horvat_impact_2011}, as
well as efforts on considering the stochastic nature of the Drake
equation \citep{glade_stochastic_2011}.
%
On this topic a number of works consider
numerical simulations \citep{forgan_evaluating_2015,
vukotic_grandeur_2016, murante_simulating_2015, forgan_numerical_2009,
forgan_galactic_2017, ramirez_new_2017}.
%
Although this approach does not rely on values that can be measured
from observations
(like the fraction of stars with planets), 
it depends on the definition of
unknown or uncertain parameters required to carry out
a the simulations.
%
\citet{grimaldi_signal_2017} considers a statistical model for the
probability of the Earth contacting other intelligent civilisations,
taking into account the finite lifetime of signal emitters, and based
on the fractional volume occupied by all signals reaching our planet.

@@

A notable example of statistical modeling is the work by
\citet{balbi_impact_2018}, who investigates the chance of
communicating civilisations making causal contact within a volume
surrounding the location of the Earth.
%
The author argues that the causal contact requirement involves mainly
the distance between civilisations, their lifespan and their times of
appearance.
%
This is important since the time the light takes to travel across the
Galaxy might be much lesser than the lifetime of the emitter.
%
\citet{balbi_impact_2018} fixes the total number of civilisations and
explores the parametric space that comprises three parameters, namely,
the distance to the Earth, the time of appearance and the lifespan of
the communicating civilization.
%
Each of these three variables are drawn from a random distribution.
%
The distribution of the distances results from a uniform model for the
positions of civilizations within the plane of the Galaxy.
%
For the distribution of the characteristic time of appearance, the
author explores exponential and truncated Gaussian distributions,
while for the distribution of the lifespans chooses an exponential
distribution.
%
It is important to consider that the estimation of the number of
communicating civilisations vary with the choice of the statistical
distribution for the time of appearance, as shown by
\citet{balbi_impact_2018}.
%
For all analysed distributions, the author concludes that, the
fraction of emitters that are listened is low if they are spreaded in
time and with limited lifetimes.
%
In another numerical approach, \citet{vukotic_astrobiological_2012}
propose a probabilistic cellular automata modelling. 
%
In this framework, a complex system is modelled by a lattice of cells
which evolve at discrete time steps, according to transition rules
that take into account the neighbour cells.
%
The authors implement this model to a network of cells which represent
life complexity on a 2-dimensional region resembling the Galactic
Habitable Zone (GHZ), an annular ring set between a minimal radius of
6 kpc and a peak radius of 10 kpc.
%
These simulations represent the spread if intelligence as an
implementation of panspermia theories.
%
Within this framework, \citet{vukotic_astrobiological_2012} also make
several Monte Carlo simulations and analyse ensemble-averaged results.
%
Their work aims at analysing the evolution of life, and although it
does not account for the network of causal contacts among
technological civilisations, it offers a tool to think SETI searches
from a novel point of view.

@@

This study is conceived as an introduction to a simple probabilistic
model and a numerical exploration of its parameter space.
%
With these tools, we address the problem of the temporal and spatial structure of
the distribution of communicating civilisations.
%
This approach does not require asumptions about, e.g., the origin of life, the development of
intelligence or the formation of habitable planets according to stellar type.
%
Instead, we assume monoparametric function families to model the appearance of
points in the disk of the Galaxy and over the time, which we call
\textit{nodes}.
%
These nodes represent the locations of ideal intelligent agents that are able
to receive and emit signals with perfect efficiency in all directions.
%
Then we analyse the network of causally connected nodes limited to have a
maximum separation representing the maximum distance a signal can travel with
an intensity above a fixed threshold.
%
A node is causally connected to other node if it is within their light cone.
%
We adopt as a definition that a light cone is the region in spacetime within
the surface generated by the light emanating from a given point in space in a
given period of time.
%
For example, two nodes separated by 100 light years which appear with a
difference in time of 80 years and last 50 years each, will be causally
connected for 30 years.
%
This connection is not, though, bidirectional.
%
In fact, the second node will see the first one for 30 years, but the first
node will never acknowledge the existence of the second.
%
This simple tought experiment exemplify the importance of the time variable to
analyse the structure of connections.
%
The model can be described with three parameters, namely, the mean separation
between the appearance of nodes, the mean lifetime of the nodes, and the
maximum separation to allow an effective causal contact.
%
The model presented here comprises the stochastic networks of constrained
causally connected nodes, together with the three parameters and additional
hypotheses.
%
Hereafter, we refer to this model as the SC3Net model, standing for
''Stochastic Constrained Causally Connected Network``.

We provide a framework to explore, through a suite of numerical simulations,
the parameter space of three unknown observables.
%
This allows to discuss possible scenarios and their consequences in terms of
the probability of making contacts.
%
The method we use for simulating a stochastic process is an approximation that
allows to study the behaviour of complex systems, by considering a sequence of
well defined discrete events.
%
The simulation is carried out by following all the variables that describe and
constitute the state of the system.
%
The evolution of the process is then
described as the set of changes in those variables. 
%
In this context, an event produces a specific change in the state that can be
triggered by random variables that encode the stochastic nature of the physical
phenomenon.
%
For example, when a new contact is produced between two entities in the
simulated galaxy, the number of active communication lines is increased by one.
%
Also, if it is the first contact for that nodes, 
then the number of communicated nodes increases by two.
%
When a new node becomes active, the sistem has an increase of one
in the number of nodes, although the number of communcations does not
necessarily change.
%
The process then involves following the changes on the state of the system, defining
the initial and final states.
%
This is done by defining a method that allows to keep track of the time
progress in steps and maintaining a list of relevant events, i.e., the events
that produce a change in the vriables of interest.
%
With this method we abandon the frequentist approach of the Drake equation to
compute the number of civilisations, providing instead its statistical
distribution.
%
More importantly, we are interested on the probability of contacts, which
depend critically on the time variable.
%
This is an exploratory analysis that aims at developing a numerical tool to
discuss the different scenarios based on statistical heuristics.
%
The approach proposed here should be considered as a compromise between the
uncertainties of the frequentist estimations and the detailed recipes required
on the numerical simulations.
%
It is worth noticing that the SC3Net model is not intended for a formal fit at this
stage due to the lack of data, but it can help to understand how unusual it
would be to actively search for intelligent signals for 50 years without
possitive detections.

This paper is organized as follows.
%
In Sec.~\ref{S_methods} we introduce the methods and discuss the candidate
distributions for the statistical aspects of the times involved in the
communication process.
%
We present our results in Sec~\ref{S_results}, with special emphasis on the
statistical distributions of the duration of causal contacts in one or both
directions and the distribution of time intervals for the waiting of the first
contact.
%
This quantities are considered as a function of the three simulation
parameters.
%
In Sec.~\ref{S_discussion} we discuss our results and future research
directions.

\begin{figure}[!t]
   %
   \centering
   %
   \includegraphics[width=0.5\textwidth]{F_scheme.pdf}
   %
   \caption{
   %
Space--time diagrams showing a schematic representation of the different stages
in the development of a constrained causal contact node (\ceti{}).
%
Panel (a) represents the region $R$ in space and time which is causally
connected to the emitter (E, left vertex), following an ``A'' type
event (``Awakening'') in which the node acquires the communication
capability.
%
The sphere of first contact (SFC) is the sphere centred on the emitter that
grows until its radius reaches the D$_{\rm max}$ distance, in which the
power of the signal would equal the detectability threshold.
%
In the Figure this sphere is represented by the left triangle of the region
$R$.
%
The surface of last contact (SLC) is another sphere that grows from a ``D''
event (``Doomsday'').
%
The region which is causally connected to the emitter is then limited by these
two spheres, and has the shape of a sphere or of a spherical shell,
depending on the time.
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b).
%
The receiver node E$_1$ can listen signals from the emitter node E$_2$, from
the ``Contact'' event (t=C$_{12}$) up to the ``Blackout'' event
(t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$. 
%
} \label{F_scheme}
%
\end{figure}

\section{Methods and working hypotheses}\label{S_methods}

Simulations are suitable tools to analyse systems that evolve with time and
involve randomness.
%
An advantage of a numerical approach is that it usually requires fewer assumptions
and simplifications, and can be applied to systems where analytical models are
hard or impossible to develop.
%
In particular, a suitable tool to model complex stochastic processes through
the changes in the state of a system is the discrete--event (hereafter, DE)
simulation approach.
%
A system described with the DE paradigm is characterised by a set of actors and
events.
%
Actors interact causally through a series of events on a timeline and process
them in chronological order \citep{ptolemaeus_system_2014,
chung_simulation_2003, ross_simulation_2012}.
%
Each event modify the variables that define the process, producing the
corresponding change in the state of the simulated system.
%
This method is well suited for the particular case of the diffusion of
intelligent signals in the galaxy and allows to explore several models easily.
%
We simulate the statistical properties of a set of points in space and time
that have a causal connection at light speed and are separated
by a maximum distance.
%
We refer to this nodes as ''Constrained Causal Contact Nodes``
(hereafter, \ceti{}).
%
We choose this generic name in order to stress the fact that in this analysis 
only the causal contact is considered, independently of any broadcasting or
lookout activity.
%
The system we propose is ideal, in the sense that it considers the special case
of a fully efficient node that emits and receives isotropically.
%
A causal contact node, considered as a broadcasting station that has the
ability to detect signals through an active search program, could have a
lesser than one efficiency factor, which is not included in the current
analysis.
%
Also, it is worth mentioning that this is a general approach, and not
necessarily a \ceti{} is the host of an intelligent civilisation.
%
It can be associated with a planet where life has developed,
became intelligent, reached the skills required to find the right communication
channel, sustained a search and established a contact.
%
Alternative message processing entities could be considered, for example
interstellar beacons where intelligent life has ceased to exist but continue with its
emission, or communication stations established by probes or left by
intelligent beings \citep[see, e.g., ][]{peters_outer_2018,
barlow_galactic_2013}.
%
In principle, these strategies could affect our results since it would be
easier to configure a cluster of \cetis{} that spread in time.
%
However, we do not consider these speculative alternatives at this point.
%
For the purpose of the present analysis, only the communication capability is
relevant, since we study the causal contacts between the locations.
%
The system is defined by a number of actors that represent \cetis{} and appear
at different instants in time, generating events that produce meaningful
changes in the variables that describe the system, i.e., in the arrangement of
\cetis{} and their network of causal contacts.
%
For example, the appearance of a new \ceti{} in a region filled with a signal
emitted by another node, will increase the number of active \cetis{} and the
number of pairs of \cetis{} in causal contact.
%
Assuming some simple hypotheses, the discrete events method can be performed
taking into account a small number of variables, which allow to analyze the
variation of the results in the SC3Net model parameter space.

@@

The temporal structure that emerges from the experimental setup implies that,
as a fundamental property of the simulations, a causal connection can be
produced without  requiring that the two nodes are active at the same time.
%
This property of the system arises as a consequence of the large spatial and
temporal scales, where a message is transmitted at the (relatively small) light
speed.
%
Although a \ceti{} could be active for a large enough period to transmit a
message at large distances, the limited power of the message and the dilution
that depends on the squared distance from the source imposes a detectability
limit.
%
As a consequence of this limitation and of their finite lifetime, considered as
the period between the acquisition and loss of communicating capacity, each
\ceti{} will fill a spherical shell region of the galaxy, limited by two
concentric spherical surfaces.
%
The leading front, or surface of first contact (SFC) grows from the central
\ceti{} until it reaches the maximum distance $D_{\rm max}$.
%
Following the end of the civilisation, there is still a region which is filled
with the emitted signals.
%
This approach has been also considered in other statistical models \citep[e.g.,
][]{smith_broadcasting_2009, grimaldi_signal_2017, Grimaldi2018}.
%
The trailing front, or surface of last contact (SLC) also grows from the
central \ceti{}, with a delay with respect to the SFC equivalent to the
lifetime of the node, and produces a spherical shell region.
%
Any other \ceti{} within this region will be in causal contact with the originating node,
even if it has disappeared before the time of contact.
%
This region will grow if the surface of first contact has not yet reached the
maximum distance $D_{\rm max}$, and will shrink after a D-type event until the
surface of last contact reaches $D_{\rm max}$, producing as a result the loss
of all signal from the central \ceti{}.
%
In our approach, we consider a model galaxy where the width of the disk is
negligible with respect to the radius of the disk.
%
In the 2D simulation only the intersection of the communicating spherical
shells with the plane of the galaxy is relevant, and produce the corresponding
circles or rings for the filled spheres or annular regions of the spherical
shells, respectively.
%
The initially growing communicating sphere is shown over space--time diagrams,
where time is represented on the horizontal direction, and space is represented
in the vertical direction.
%
In the Fig.~\ref{F_scheme} the two emitters in panel (b), E$_1$ and E$_2$,
reach each other at different times.
%
The time span for E$_i$ is (A$_i$ , D$_i$) , for $i = {1, 2}$.
%
Emitter $i$ can listen to emitter $j$ between C$_{ij}$ and B$_{ij}$.
%
The type and length of causal contact in both directions depend on the distance
and time lag between the awakening events, the maximum distance that a signal
can reach and the time period in which each emitter is active.

@@

   $\tau_{a}$ & Mean temporal separation between consecutive
   awakenings & 4 kyr & 200 kyr & 50\\ 
   $\tau_{s}$ & Mean lifetime of a \ceti{}
   & 10 kyr & 500 kyr& 50\\ 
   D$_{max}$ & Maximum reach of a message  &  \multicolumn{2}{l}{3 kpc, 12 kpc, 25 kpc} & 3\\
\hline
